Article title

### 2017-07-31 paper-webserver.tex
ADD: more revisions

% costs of bacterial infections
Foodborne ilnesses are increasingly difficult to track; ingrediants are sourced globally and contamination can occur at any point in the supply-chain.
The resulting consumer infections translate into actual costs: in Canada, an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
In the United States, this figure is 55961 hospitalizations and 1351 deaths per year \cite{scallan2011foodborne}.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.
% surveillance
Public-health agencies routinely monitor emerging threats, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, a variety of bioinformatic approaches are now based on whole-genome sequencing (WGS).

% traditional reference lab techniques
To begin analyzing genomes, researchers designate particular strains of bacterial isolates using identifiers known as subtypes.
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, pulsed-field gel electrophoresis (PFGE), or Sanger sequencing, amongst others \cite{ronholm2016navigating}.
PFGE is unique amongst the set; restriction enzymes are used to produce DNA fragments visible as specific banding patterns.
Public health agencies use PFGE to track outbreaks by comparing band patterns of bacterial isolates \cite{kozak2013foodborne}.
However, as WGS rarely produces complete circular genomes and, in PFGE, cutting by a restriction enzyme is dependent on protein structure, there is not a bioinformatic-analog to PFGE.
Serotyping uses antibodies to detect the presence of specific cell surface antigens.
Sanger sequencing is historically used to examine differences in housekeeping genes which are required to maintain cell function, an approach known as multi-locus sequence typing (MLST).
Unlike PFGE, both serotyping and MLST have WGS-based computational methods which mimick results of the wet-lab techniques.
Software such as VirulenceFinder \cite{joensen2014real} compares a pre-collected set of reference VF gene sequences against a WGS result.
Similar approaches are applied to predict AMR genes \cite{mcarthur2013comprehensive}, O-type, and H-type \cite{whiteside2016superphy}.

% potential of WGS
WGS expands on traditional methods: due to the inclusion of the entire genome, WGS-based methods offer better resolution and context for evaluating the spread of outbreaks \cite{ronholm2016navigating}.
In the case of laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods, such as core genome MLST (cgMLST), attempt to cover every common gene.
We can also detect the presence of every known virulence factor (VF) or antimicrobial resistance (AMR) gene in a WGS result.
Furthermore, using WGS we can study the relations between genomes, changes in the VF or AMR genes present over time, and mutations in gene \cite{ronholm2016navigating}.
Entire datasets and results can also be shared in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods only examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, more target sequences are identified using WGS; if we take a tradition approach to making threat assessments by looking for known VFs or AMR genes, the large number of results make it difficult to guage risk \cite{ronholm2016navigating}.
The effectiveness of WGS-based methods also rely on the accuracy and breadth of reference datasets.
Agencies must also weigh the cost of developing people and knowledge in bioinformatic approaches, and laboratory based phenotypic tests remain inexpensive \cite{koser2012routine}.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc} without employing software engineering principles \cite{de2015trends}.
Tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
As WGS-technology expanded the availability of data, systems engineering principles became important since code failure could quickly disrupt entire bioinformatics pipelines \cite{schatz2015biological}.
The field also lags behind other big data industries: bioinformatics software still struggles to perform computations in real-time, stores data in inflexible databases, doesn't support live visualizations \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

% Solution via software engineering
Efforts such as the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} attempt to alleviate data concerns by curating a set of known AMR genes.
In addition, agencies such as the Food and Drug Administration (FDA), which hosts GenomeTrakr \cite{allard2016practical}, the National Center for Biotechnology Information (NCBI), which hosts GenBank \cite{doi:10.1093/nar/gks1195}, and the Center for Disease Control and Prevention (CDC), which hosts PulseNet \cite{swaminathan2001pulsenet}, provide services to share genome sequences and associated metadata either openly or amongst member institutions.
Recent efforts have focused on providing modern tools for quality control of reference genomes \cite{parks2015checkm}, particularly amongst human-genome projects \cite{lee2017ngscheckmate}.

% rewrite of bigdata approach (BIGGER bigdata)
While efforts such as CARD as important for replicating traditional results and for historical context, WGS can also be used to develop new methods using the larger datasets available.
To study entire genomes, we need methods of creating reference data encompassing all genes, both core and accessory, in a given phlogeny.
Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, chewBBACA \url{https://github.com/mickaelsilva/chewBBACA}, and Graphtyper \cite{Eggertsson148403} are freely available to compute these pan-genomic regions.
These software approaches must be scalable to significantly larger datasets than those found in wet-lab approaches.
Also, for uptake by front-line clinicians and laboratories, WGS solutions should perform faster; for example, AMR genes should be determined faster than the typical 18-24 hours required in a reference laboratory \cite{koser2012routine}.
Storing results in a tiered structure, with rapid access available for retrieval and visualization while maintaining long-term data storage \cite{schatz2015biological}, would enable developers to address the current limitations \cite{de2015trends} of bioinformatics software.

@@

% this study
Here we present Spfy, a complete rewrite of Superphy to put a graph databases at the core.
Graph databases focus on describing the relationships between different data, and is one of the emerging \cite{de2015trends} database types being used for biological data.
Spfy takes a modular approach to tool integration and distributes analyses over task queues, thereby allowing modules to run in parallel and address code failures.
Serotyping, VF, AMR predictions are computed within minutes, and results are efficiently stored within the graph.
Furthermore, large-scale analyses, such as epidemiological associations between specific genotypes, biomarkers, host, source, and other metadata, and statistically significance testing of genome markers for user-defined groups, are provided in real-time.
We also containerize services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Docker integration ensures that software dependencies, which typically must be manually installed \cite{doi:10.1093/bioinformatics/btu153,laing2010pan,inouye2014srst2}, are handled automatically.
Furthermore, 
These changes enable us to provide real-time genome analysis and database storage of results for upwards of 5000 genomes.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.

>>>>

% costs of bacterial infections
Foodborne ilnesses are increasingly difficult to track; ingrediants are globally sourced and contamination can occur at any point in the supply-chain.
The resulting consumer infections translate into actual costs: in Canada, an estimated 4000 hospitalizations and 105 deaths per year are thought to be the result of various pathogens \cite{thomas2015estimates}.
In the United States, this figure is 55961 hospitalizations and 1351 deaths per year \cite{scallan2011foodborne}.
Futhermore, the associated economic costs have been known for decades \cite{todd1989costs}.
% surveillance
Public-health agencies routinely monitor emerging threats, of which \textit{Salmonella enterica} and \textit{Escherichia coli} are amongst the two most common \cite{kozak2013foodborne}.
In \textit{Salmonella enterica} \cite{bell2016recent}, \textit{Escherichia coli} \cite{fratamico2016advances}, and others pathogens \cite{ronholm2016navigating}, a variety of bioinformatic approaches to surveillance are now based on whole-genome sequencing (WGS).

% traditional reference lab techniques
To begin analyzing genomes, researchers designate particular strains of bacterial isolates using identifiers known as subtypes.
Traditional reference laboratory techniques handled subtyping through methods such as serotyping, pulsed-field gel electrophoresis (PFGE), or Sanger sequencing, amongst others \cite{ronholm2016navigating}.
PFGE is unique amongst the set; restriction enzymes are used to produce DNA fragments visible as specific banding patterns.
Public health agencies use PFGE to track outbreaks by comparing band patterns of isolates \cite{kozak2013foodborne}.
However, as WGS rarely produces complete circular genomes and, in PFGE, cutting by a restriction enzyme is dependent on protein structure, there is no bioinformatic-analog to PFGE.
Unlike PFGE, both serotyping and MLST have WGS-based computational methods which mimick results of the wet-lab techniques.
Serotyping uses antibodies to detect the presence of specific cell surface antigens and multi-locus sequence typing (MLST), using Sanger sequencing, examines differences in housekeeping genes which are required to maintain cell function.
By comparing a pre-collected set of reference VF gene sequences against a WGS result, we can predict wet-lab serotyping results \cite{whiteside2016superphy}, Virulence Factor (VF) results \cite{joensen2014real}, and antimicrobial resistance (AMR) results \cite{mcarthur2013comprehensive}.

% potential of WGS
WGS also expands on traditional methods: due to the inclusion of the entire genome, WGS-based methods offer better resolution and context for evaluating the spread of outbreaks \cite{ronholm2016navigating}.
In the case of laboratory-based MLST, instead of examining selected single-nucleotide polymorphisms (SNPs) in a subset of housekeeping genes, newer WGS-based methods, such as core genome MLST (cgMLST), attempt to cover every common gene.
Instead of detecting only a subset of VFs and AMR genes, we can detect every known sequence in a WGS result.
Furthermore, using WGS we can study the relations between genomes, changes in the VF or AMR genes present over time, and mutations in genes \cite{ronholm2016navigating}.
Entire datasets and results can also be shared in real-time \cite{ronholm2016navigating}.

% Drawbacks - genetics
However, these WGS-based methods only examine the underlying genetics, not the expressed phenotype; the presence of a particular gene sequence does not necessitate expression.
Correspondingly, more target sequences are identified using WGS; if we take a tradition approach to making threat assessments by looking for known VFs or AMR genes, the large number of results make it difficult to guage risk \cite{ronholm2016navigating}.
The effectiveness of WGS-based methods also rely on the accuracy and breadth of reference datasets and agencies must weigh the cost of developing expertise in bioinformatic approaches, while laboratory based phenotypic tests remain inexpensive \cite{koser2012routine}.

% drawbacks - big data
Traditionaly, bioinformatics software was developed \textit{ad hoc} without employing software engineering principles \cite{de2015trends}.
Tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
Since WGS-technology expanded the availability of data, systems engineering principles are important since code failure must be handled throughout the entire bioinformatics pipeline \cite{schatz2015biological}.
The field also lags behind other big-data industries: bioinformatics software still struggles to perform computations in real-time, stores data in inflexible databases, doesn't support live visualizations \cite{de2015trends}, and lacks a common application programming interface (API) linking resources \cite{swaminathan2016review}.

% Solution via software engineering
Efforts such as the Comprehensive Antibiotic Resistance Database (CARD) \cite{mcarthur2013comprehensive} attempt to alleviate data concerns by curating a set of known AMR genes.
In addition, agencies such as the Food and Drug Administration (FDA), which hosts GenomeTrakr \cite{allard2016practical}, the National Center for Biotechnology Information (NCBI), which hosts GenBank \cite{doi:10.1093/nar/gks1195}, and the Center for Disease Control and Prevention (CDC), which hosts PulseNet \cite{swaminathan2001pulsenet}, provide services to share genome sequences and associated metadata either openly or amongst member institutions.
Recent efforts have focused on improving the quality of shared reference genomes \cite{parks2015checkm}, particularly amongst human-genome projects \cite{lee2017ngscheckmate}.

% rewrite of bigdata approach (BIGGER bigdata)
While efforts such as CARD as important for replicating traditional results and for historical context, WGS can also be used to develop new methods using the larger datasets available.
To study entire genomes, we need methods of creating reference data encompassing all genes in a given phlogeny.
Panseq \cite{laing2010pan}, Roary \cite{page2015roary}, chewBBACA \url{https://github.com/mickaelsilva/chewBBACA}, and Graphtyper \cite{Eggertsson148403} are freely available to compute these pan-genomic regions.
These software approaches must be scalable to larger datasets than those found in wet-lab approaches.
Also, for uptake by front-line clinicians and laboratories, WGS solutions should perform faster; for example, AMR gene prediction should take less than the typical 18-24 hours required in a reference laboratory \cite{koser2012routine}.
Storing results in a tiered structure, with rapid access available for retrieval and visualization while maintaining long-term data storage \cite{schatz2015biological}, would enable developers to address the current interactivity limitations \cite{de2015trends} of bioinformatics software.

@@

% this study
Here we present Spfy, a complete rewrite of Superphy.
Spfy is backed by a graph database which our analyses modules rely on.
Graph databases focus on describing the relationships between different data, and is one of the emerging \cite{de2015trends} database types being used for biological data.
Spfy takes a modular approach to tool integration and distributes analyses over task queues, thereby allowing modules to run in parallel and address code failures.
Serotyping, VF, AMR predictions are computed within minutes, and results are efficiently stored within the graph.
Furthermore, large-scale analyses, such as epidemiological associations between specific genotypes, biomarkers, host, source, and other metadata, and statistically significance testing of genome markers for user-defined groups, are provided in real-time.
We also containerize services using Docker, a modern approach found in many web servers and proposed to mitigate the complexities of bioinformatic pipelines \cite{di2015impact}.
Docker integration ensures that software dependencies, which typically must be manually installed \cite{doi:10.1093/bioinformatics/btu153,laing2010pan,inouye2014srst2}, are handled automatically.
These changes enable us to provide real-time genome analysis and database storage of results for upwards of 5000 genomes.
All services are publicly available at \url{https://lfz.corefacility.ca/superphy/spfy/}.
